---
title: "Code Sample- Yearly Spatial Analysis of PM2.5 Distribution in Guatemala"
author: "Chiara Brust"
date: "`r Sys.Date()`"
output:
  html_document: default
  pdf_document: default
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


## Load libraries
```{r}

library(sf)
library(raster)
library(tidyverse)
library(ggplot2)
library(readxl)
library(tmap)

```



## Import Data- Guatemala shapefile
```{r}
guat_shp <- st_read("Data/gtm_admbnda_adm1.shp")


# test <- guat_shp %>% filter(ADM1_ES == "Alta Verapaz")
# rm(test)

str(guat_shp)

```



## Set coordinate reference system
crs for WGS 84 4326- (https://spatialreference.org/ref/epsg/wgs-84/)
```{r}

# Set coordinate reference system
guat_shp  <- guat_shp %>%
  st_set_crs(4326)


# check crs
st_crs(guat_shp)

```



## Check out the shapefile
```{r}
plot(guat_shp)

tm_shape(guat_shp)+
  tm_polygons() +
  tm_compass(
    type= "arrow",
    position= c("right", "top")
  ) +
  tm_scale_bar()

```


## Rename the `Department` variable
```{r}
guat_shp <- guat_shp %>%
  rename(Department = ADM1_ES)
```


## Import- Yearly PM2.5 Data
Then rename the `mid-long` and `mid-lat` coordinate columns
```{r}
guat_data <- read_excel("Data/wustl_guatemala_pm25_0p1grid1998-2020.xlsx")

str(guat_data)

# Renaming PM2.5 Data coordinate columns
guat_data <- rename(guat_data, "X" = "mid-long", "Y" = "mid-lat")

```



## Import- Coordinate and Department data
```{r}
guat_grids_csv <- read.csv("Data/grids_guatemala.csv")

guat_grids_dept <-read.csv("Data/grids_guatemala_intersect.csv")

str(guat_grids_csv)
str(guat_grids_dept)
```



## Join departments with coordinates
```{r}
coord_dept_join <- right_join(guat_grids_csv, guat_grids_dept, by = c("Maille", "Maille_X", "Maille_Y"))


# Rearrange columns
coord_dept_join1 <- coord_dept_join %>%
  arrange(ADM1_ES) %>%
  select(ADM1_ES, X:ADM1_PCODE, everything()) %>%
  rename(Department = ADM1_ES)

```



## Joining the PM2.5 data with the coordinates and departments data
I'm using `right_join` instead of `full_join` because we have some PM2.5 data outside of Guatemala
```{r}
PM_coord <- right_join(guat_data, coord_dept_join1, by = c("X", "Y"))


# Rearranging the data and checking out the variables
PM_coord <- PM_coord %>% 
  arrange(Department) %>%
  select(Department, everything())

str(PM_coord)
```



## Pivoting `Year` and `PM2.5` then rearranging
I also removed the `Y` in front of each year and converted that variable into class `numeric`
```{r}
PM_pivot <- PM_coord %>%
  pivot_longer(4:26, names_to= "Year", values_to= "pm25") %>%
  mutate(Year = str_sub(Year, 2)) %>% 
  mutate(Year= as.numeric(Year)) %>% 
  select(Department, Year, pm25, everything())

str(PM_pivot)
```




# The following is a test plot of only one year's data
## Get only `x` (longitude), `y` (latitude), and `pm25` from `PM_pivot` for one year (test= `1998`)
```{r}
test_pivot_filter <- PM_pivot %>% 
  filter(Year== 1998) %>% 
  dplyr::select(4:5, 3)

str(test_pivot_filter)
```


## Converting `test_pivot_filter` into a `raster` to mask it with the `guat_shp`
I then convert it back into a `dataframe` for use in the subsequent plot
```{r}
# Create spatial points data frame from my dataframe
spg <- test_pivot_filter
coordinates(spg) <- ~ X + Y
# Coerce to SpatialPixelsDataFrame
gridded(spg) <- TRUE
# Coerce to raster
rasterDF <- raster(spg)
# Crop
plot(rasterDF)
rasterDF_crop <- crop(rasterDF, extent(guat_shp))
plot(rasterDF_crop)
# Mask
rasterDF_masked <- mask(rasterDF_crop, guat_shp)
plot(rasterDF_masked)

# Convert back to DF
df_masked <- as.data.frame(rasterDF_masked,xy=TRUE)


str(df_masked)
```



# Test Plot- `1998` air pollution
```{r}
plot98_yearly <- ggplot() +
  geom_raster(data = df_masked, aes(x = x, y = y, fill=pm25), interpolate = TRUE) +
  scale_fill_continuous(low = "white",
                        high = "red",
                        na.value = "transparent",
                        limits= c(0, max(df_masked$pm25))) +
  geom_sf(data= guat_shp, fill= NA, linewidth= .5)+
  labs(title=(expression(paste("Average ", PM[2.5], " in Guatemala in 1998"))))+
  labs(fill=expression("PM"[2.5] (mu*g/m^3)))+
  theme(plot.title = element_text(hjust = 0.5))

plot(plot98_yearly)

```










